---
title: "Data frames & Subsetting"
author: "Tim Dennis"
date: "January 7, 2016"
output: html_document
---

#### Data frames & reading in data (20m)

Become familiar with data frames & able to read data into R.

* similar to matricies but each column can be different atomic type
* tabular data spreadsheets -rectangular 
* data frames are lists under the hood
* can be crated manually

```{r}
df <- data.frame(id = c('a', 'b', 'c', 'd', 'e', 'f'), x = 1:6, y = c(214:219))
df
length(df) #tells you the number of columns not rows
nrow(df) #gives you num rows
```

* add column to data frame with cbind()

```{r}
df <- cbind(df, 6:1)
df
```
* note R gives names to the column, we can change this using `names()`
* `names(df)[4] <- 'SixToOne'`
* we can do this when adding a column

```{r}
df <- cbind(df, caps=LETTERS[1:6])
df
```

* note: LETTERS and letters are built-in constants

* add a row using rbind 
* ` <- rbind(df, list("g", 11, 42, 0, "G"))df`
* we add as a list b/c we have multiple types across columns
* what is this telling us? 
* trying to append 'g' and 'G' as factor levels, why?
* `class(df)` - `str(df)` 
* when we used the data.frame() function R automatically made the 1st & last col into factor & there are no levels for 'g' and "G", so adding fails
* `df` - row was added but those two are NAs
* workarounds: 
	- convert the factor columns into characters - but we lose the factors
	- add factor levels to accommodate the new additions. if we want to preserve factors this is what we should do
	
```{r}
# convert col #1 to character
df$id <- as.character(df$id)  # convert to character
class(df$id)
```

```{r}
levels(df$caps) <- c(levels(df$caps), 'G') # add a factor level
class(df$caps)
```

* let's add the row again

```{r}
df <- rbind(df, list("g", 11, 42, 0, 'G'))
tail(df, n=3)
```

* we should delete teh row with the NAs
* multiple ways to delete rows

```{r}
df[-7, ]  # The minus sign tells R to delete the row
df[complete.cases(df), ]  # Return `TRUE` when no missing values
na.omit(df)  # Another function for the same purpose
df <- na.omit(df) # overwrite to save

```

* combinding data frames
* row-bind data.frames together
* `rbind(df,df) #r is making sure rownames are unique`
* restor sequential numbering by setting rownames to NULL

```{r}
df2 <- rbind(df, df)
rownames(df2) <- NULL
df2
```

*  READing in data - we have data in a 'data/' folder, let's look at it in a text editor `sublime gap` 

```{r}
#using read.table(), sep ','
gapminder <- read.table(
  file="data/gapminder-FiveYearData.csv",
  header=TRUE, sep=","
)
head(gapminder)
```

```{r}
# Here is the read.csv version of the above command
gapminder <- read.csv(
  file="data/gapminder-FiveYearData.csv"
)
head(gapminder)

```

* another type of file tab-separated format, use sep="\t" for these
* you can also read in files from the web by using url 
* for excel use package like xlsx 
* Make reproducible, let's create a R script File > new file > R script  - show how to run using source (mention you will save all commands to dropbox)

**Using gapminder data**

* remember functions for inspecting data structures.

```{r eval=FALSE}
class() # what is the data structure?
typeof() # what is its atomic type?
length() # how long is it? What about two dimensional objects?
attributes() # does it have any metadata?
str() # A full summary of the entire object
dim() # Dimensions of the object - also try nrow(), ncol()
```

```{r}
#apply to gapminder
typeof(gapminder) #data frames are lists
class(gapminder) #class of df
head(gapminder)
typeof(gapminder$year) #looking to see what type year is
#Can anyone guess what we should expect the type of the continent column to be?
typeof(gapminder$continent) #integer why?
class(gapminder$continent) #factors are stored as integers

```

* default behavior of R to treat text columns as factors when reading in
* this is b/c text columns often represent categorical data, which need to be factors for statistical modeling 
* not obvious can be confusing
* we can disable behavior

```{r}
options(stringsAsFactors=FALSE)
gapminder <- read.table(
  file="data/gapminder-FiveYearData.csv", header=TRUE, sep=","
)
```

* remember that you'll need to convert to factors when needed
* can be good b/c it forces you to think about the questions you are asking and maybe easier to order categories
* `str(gapminder)`
* read str
* we can modify column/row names
* `colnames(gapminder)`

```{r}
copy <- gapminder
colnames(copy) <- letters[1:6]
head(copy, n=3)

```

* rownames: `rownames(gapminder)[1:20]`
* sequential row nums
* attributes will give row and col names
* dimnames will give just rownames
* `str(dimnames(gapminder))`

**understanding lists in functional outputs**
* run a basic linear regression with gapminder

```{r}
# What is the relationship between life expectancy and year?
reg <- lm(lifeExp ~ year, data=gapminder)
```

* lm estimates linear statistical models
* `a~b` meaning a, the dependent (or response) var, is a function of b, the independent variable.
* we tell lm to use gapminder so it knows where lifeExp and year are
* output `reg` - not too much
* `str(reg)` - there's a great deal stored in nested lists
* `summary(reg)` - we can look at summary - life expectancy has slowly been increasing over time, sig positive association

#### Subsetting data (25m)
To be able to subset vectors, factors, matrices, lists, and data frames

* six different ways we can subset any kind of object
* 3 diff subsetting operators
* start with an atomic vector

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```

**Accessing elements by indicies**
* extract element by index
* [] means give me the xth element
* `x[1]`
* `x[4]`
* we can ask for multiple elements
* `x[c(1,3)]`
* or slices of a vector `x[1:4]`
   * x[1:4] is equivalent to x[c(1,2,3,4)].
* we can ask for repeating elements `x[c(1,1,3)]`
* if we ask for element out of scope we get NA `x[6]`
* `x[0]` get us a empty vector

**Skipping and removing elements** 

* using negative number will return every element except for that one
* `x[-2]` 
* skip multiple elements `x[c(-1,-5)] # or x[-c(1,5)]`
* `x[-1:3]` creates an error b/c its reading as c(-1,0,1,2,3)
    - needs to use parentheses so order is right 
    - `x[-(1:3)]`
* to remove elements from a vector we need to assign the result back over the var

```{r}
x <- x[-4]
x
```

**Subsetting by name**

* extract elements by using their name
* `x[c("a","c")]` 
    - usuall more reliable way to subset objects, index changes more than names
    - unfort can't skip or remove as easily
* `x[-which(names(x)=="a")]` # which function returns all true elements of its argument, which is negated 
* let's step through this: 

```{r}
names(x) == "a" # the condition applied to vector x, only a true
which(names(x)=="a") # which converts this to a index
x[-which(names(x)=="a")] # using the "-" will remove the element at that index
```

* skipping multiple named indicies uses %in%

* `x[-which(names(x) %in% c("a", "c"))]`
* So why can’t we use == like above? 
* Because '==' expects objects of the same length and if not, will start repeating the shorter object
* Called recycling rule 

**R recycling rule** 

```{r eval=F}
names(x) == c('a', 'c') #warnings 
#== works slightly differently than %in%. It will compare each element of its left argument to the corresponding element of its right argument.
# R recycles the shorter vector in a equality comparison

c("a", "b", "c", "e")  # names of x
   |    |    |    |    # The elements == is comparing
c("a", "c")

c("a", "b", "c", "e")  # names of x
   |    |    |    |    
c("a", "c", "a", "c") # The elements == is comparing
```

*  Non-unique names are allowable for multiple elements in a vector 

```{r}
x <- 1:3
x
names(x) <- c('a', 'a', 'a')  
x
x['a']  # only returns first value
x[which(names(x) == 'a')]  # returns all three values
```

* Getting help for operators; 
    - `help("%in%")` or `?"%in%"`

**Subsetting through other logical operators**

* we can subset through logical operations
* `x[c(TRUE, TRUE, FALSE, FALSE)]`
* Note that in this case, the logical vector is also recycled to the length of the vector we’re subsetting!
* `x[c(TRUE, FALSE)]`
* Since comparison operators evaluate to logical vectors, we can also use them to succinctly subset vectors:
    - `x[x > 7]`

There are many situations in which you will wish to combine multiple conditions. To do so several logical operations exist in R:

* | logical OR: returns TRUE, if either the left or right are TRUE.
* & logical AND: returns TRUE if both the left and right are TRUE
* ! logical NOT: converts TRUE to FALSE and FALSE to TRUE
* && and || compare the individual elements of two vectors. Recycling rules also apply here.

**Handling special values**

At some point you will encounter functions in R which cannot handle missing, infinite, or undefined data.

special functions to deal with this: 

* is.na will return all positions in a vector, matrix, or data.frame containing NA.
* likewise, is.nan, and is.infinite will do the same for NaN and Inf. 
* is.finite will return all positions in a vector, matrix, or data.frame that do not contain NA, NaN or Inf.
* na.omit will filter out all missing values from a vector

**factor subsetting**

Factor subsetting works the same way as vector subsetting.

```{r}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
f[f %in% c("b", "c")]
f[1:3]

```
* An important note is that skipping elements will not remove the level even if no more of that category exists in the factor:
    - f[-3]

**Matrix subsetting**

Matrices are also subsetted using the [ function. In this case it takes two arguments: the first applying to the rows, the second to its columns:

```{r}
set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]
```
* You can leave the first or second arguments blank to retrieve all the rows or columns respectively:

```{r}
m[, c(3,4)] #column
m[3,] #if grabbing 1 row, R will convert to vector
m[3, , drop=FALSE] # keep as a matrix specify a third argument; drop = FALSE:
#m[, c(3,6)] #will throw error if out of range
```

* because matricies are really vectors, you can just use single indexing
    - m[5] #not very useful
* matricies are populated by column-major format by default, elements are arraged column wise

```{r}
matrix(1:6, nrow=2, ncol=3)
matrix(1:6, nrow=2, ncol=3, byrow=TRUE) #populate by row use  byrow=TRUE

```

**List subsetting**

Three functions to subset lists [, [[, $

* Using [ will always return a list. If you want to subset a list, but not extract an element, then you will likely use [.

```{r}
xlist <- list(a = "Software Carpentry", b = 1:10, data = head(iris))
xlist[1]
```

* we can subset elements teh same way as atomic vector
* xlist[1:2]
* to get at individual elements of a list you need to use [[
    - `xlist[[1]]`
    - but you cant extract more than one element at once `xlist[[1:2]]`
    - nor skip elements `xlist[[-1]]`
    - but you can use names to subset and extract `xlist[['a']]`
    - $ function extracts by name xlist$data

**Data frame**

Data frames are lists under the hood, so similar methods apply.
[ will act the same way as for lists, where each list is a column

```{r}
head(gapminder[3])
head(gapminder[["lifeExp"]]) # [[ will act to extract a single column
head(gapminder$year) #$ provides shorthand to extract columns by name
#With two arguments, [ behaves the same way as for matrices:
gapminder[1:3,]
gapminder[3,] #is a data frame b/c of the mixed types
```
