---
title: "Solutions to Week 3 In-class Problems"
author: "Hyeonsu Kang & Tim Dennis"
date: "January 22, 2016"
output: 
  html_document:
    toc: true
    theme: united
---

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggvis)
```
## 1.1 Data management with dplyr 

### 1. Read in the data 

```{r}
gapminder <- read.csv("https://goo.gl/BtBnPg", header = T, stringsAsFactors = F)
```

```{r include=FALSE}
gdp_byyear <- gapminder %>%
  group_by(year) %>%
  summarize(mean_gdpPercap=mean(gdpPercap))
```

### 2. What is the (absolute) difference between the average GDP per capita of Americas and Asia in year 2002?
```{r}
gdp_bycontinents_byyear <- gapminder %>%
  group_by(continent,year) %>%
  summarize(mean_gdpPercap=mean(gdpPercap))

```
```{r}
yr2002GdpPercapConti <- gdp_bycontinents_byyear %>% 
  filter(year==2002)
yr2002AsiaGdpPercap <- yr2002GdpPercapConti %>% 
  filter(continent=="Asia")
yr2002AmericasGdpPercap <- yr2002GdpPercapConti %>% 
  filter(continent=="Americas")
abs(yr2002AsiaGdpPercap$mean_gdpPercap - yr2002AmericasGdpPercap$mean_gdpPercap)  

```

* Another approach to do above, but using `dplyr` for all but the final step. 

```{r}
gdp_byasiaamer_byyear <- gapminder %>%
  group_by(continent,year) %>%
  filter(continent=="Asia" | continent == "Americas", year == "2002") %>%
  summarize(mean_gdpPercap=mean(gdpPercap))

abs(diff(gdp_byasiaamer_byyear$mean_gdpPercap))
```


### 3. Which continent had the highest GDP per capita in year 2002?

```{r}
yr2002ContiGdpPercap <- gdp_bycontinents_byyear %>% 
  filter(year==2002) 
```

```{r}
yr2002ContiGdpPercap$continent[which.max(yr2002ContiGdpPercap$mean_gdpPercap)]

```


* Another approach without the intermediary subsetted data frames. 
* When trying to create a ranking, you can use dplyr's `arrange` function, but it will behave oddly on grouped data. We need to ungroup to get arrange to work right. Try commenting the `ungroup()` line to see how it changes.
* We use `top_n()` to return top row, so we get a 1 row ordered data frame
* Can do `which.max()` outside of `%>>%` stanza 

```{r}
g2 <- gapminder %>%
  group_by(continent,year) %>%
  filter(year==2002) %>% 
  summarize(mean_gdpPercap=mean(gdpPercap)) %>% 
  ungroup() %>% 
  arrange(desc(mean_gdpPercap)) %>% #orders the data frame in descending order using arrange & desc
  top_n(1) 

head(g2) #top_n grabs the top n rows. 
g2$continent[which.max(g2$mean_gdpPercap)] #we can pull out the which.min as well
#g2$continent[which.min(g2$mean_gdpPercap)] #this would work if we hadn't used the top_n() function
```

### 4. Which continent had the lowest GDP per capita in year 2002?

```{r}
yr2002ContiGdpPercap$continent[which.min(yr2002ContiGdpPercap$mean_gdpPercap)]

```

### 5. Which country had the highest GDP per capita in year 1967?

```{r}
gdp_bycountries_byyear <- gapminder %>%
  group_by(country, year) %>%
  summarize(gdpPercap)
```


```{r}
yr1967CountGdpPercap <- gdp_bycountries_byyear %>% 
  filter(year==1967)

yr1967CountGdpPercap$country[which.max(yr1967CountGdpPercap$gdpPercap)]

yr1967CountGdpPercap$gdpPercap[which.max(yr1967CountGdpPercap$gdpPercap)]


```

* Another example of going from gapminder (not saving as a new data frame) to the subsetted data frame to the one row contaning the answer. You could save this as a df and then access use `my_dataframe[1,1]` to get the answer.

```{r}
gapminder %>%
  group_by(country, year) %>%
  summarize(gdpPercap) %>% 
  filter(year == 1967) %>% 
  ungroup() %>% 
  slice(which.max(gdpPercap))
 
```

### 6. Which country had the highest life expectancy in year 2002 and what was the value of it?

```{r}
lifeExp_bycountries_byyear <- gapminder %>%
  group_by(country, year) %>%
  summarize(lifeExp)
```

  
```{r}
yr2002CountLifeExp <- lifeExp_bycountries_byyear %>% 
  filter(year==2002)
```


* the name of the country

```{r}
yr2002CountLifeExp$country[which.max(yr2002CountLifeExp$lifeExp)]

```

* the value of the life expectancy

```{r}
yr2002CountLifeExp$lifeExp[which.max(yr2002CountLifeExp$lifeExp)]

```

### 7. What does the distribution of the log of GDP per capita look like in year 1967? 

```{r}
hist(log(yr1967CountGdpPercap$gdpPercap), plot=T, main="Histogram of Each Country's log(GDP per capita in 1967)", xlab="log(GDP per capita in 1967)")

```

### 8. What does the distribution of the life expectancy in year 2002 look like?

```{r}
hist(yr2002CountLifeExp$lifeExp, plot=T, main="Histogram of Each Country's Life Expectancy in 2002", xlab="Life Expectancy in 2002")
```

* another way using a viz package that works with %>%

```{r}
lifeExp_bycountries_byyear %>% 
  filter(year==2002) %>%
  ggvis(~lifeExp) %>%
  layer_histograms()
```

### 9. Add a log(GDP per capita in 1967) column to the data frame that contains GDP per capita by countries in 1967.

```{r}
yr1967CountGdpPercapLog <- yr1967CountGdpPercap %>% 
  mutate(logGDP=log(gdpPercap))

head(yr1967CountGdpPercapLog)
```

### 10. Graph (scatter plot) x:Log(GDP per capita in 1967), y:Life Expectancy in 2002

```{r}
yr1967GDPyr2002LifeExp <- merge(yr1967CountGdpPercapLog, yr2002CountLifeExp, by.x="country", by.y="country")
head(yr1967GDPyr2002LifeExp)
```

```{r}
ggplot(data=yr1967GDPyr2002LifeExp, aes(x=yr1967GDPyr2002LifeExp$logGDP, y=yr1967GDPyr2002LifeExp$lifeExp)) + 
  geom_point() + 
  xlab("log(GDP per capita 1967)") + 
  ylab("Life Expectancy in 2002") + 
  ggtitle("Log GDP per capita (1967) - Life Expectancy (2002) Graph") + 
  geom_text(aes(label=yr1967GDPyr2002LifeExp$country), size=3)
```

### 11. Graph a scatter plot with an x axis of \log(GDP per capita per continent in 1967)", and a y axis of \Life expectancy per continent in 1967".

```{r}
lifeExp_bycontinents_byyear <- gapminder %>%
  group_by(continent,year) %>%
  summarize(mean_lifeExp=mean(lifeExp))
```

```{r}
yr1967ContiGdpPercap <- gdp_bycontinents_byyear %>% 
  filter(year==1967) 
yr1967ContiLifeExp <- lifeExp_bycontinents_byyear %>% 
  filter(year==1967)

```

```{r}

yr1967ContiLogGdpPercap <- yr1967ContiGdpPercap %>% 
  mutate(logGDP=log(mean_gdpPercap))

```

```{r}
yr1967GDPyr1967LifeExp <- merge(yr1967ContiLogGdpPercap, yr1967ContiLifeExp, by.x="continent", by.y="continent")
ggplot(data=yr1967GDPyr1967LifeExp, aes(x=yr1967GDPyr1967LifeExp$logGDP, y=yr1967GDPyr1967LifeExp$mean_lifeExp)) + geom_point() + 
  xlab("log(GDP per capita per continent 1967") + 
  ylab("Life Expectancy per continent in 1967") + 
  ggtitle("Log GDP per capita and Life Expectancy per continent (1967)")

```

### 12. Graph (line plot) x: year, y: Log(GDP per capita per continent)

```{r}
LogGdp_bycontinents_byyear <- gdp_bycontinents_byyear %>%
  mutate(logGDP=log(mean_gdpPercap))
ggplot(data=LogGdp_bycontinents_byyear, aes(x=LogGdp_bycontinents_byyear$year, y=LogGdp_bycontinents_byyear$logGDP, color=continent)) + 
  geom_line() +
  xlab("Year") +
  ylab("Log GDP per capita per continent") +
  ggtitle("The trend of GDP per capita in each continent")
```

* Below, we use **ggvis** instead of **ggplot2** to create a line graph of year by a log of GDP per continent. 
* **ggvis** is under active development and doesn't have all the features of ggplot2. For instance, there isn'ta plot title (ggtitle)
* Notice how we just use the `log()` function inside the `ggvis` function instead of using `mutate()` -- we could also do this transformation on the fly in ggplot2 as well. The choice of transforming in the graphing function or creating a new variable in the data frame depends on if you need that variable to persist for longer than a plot. If not, do the tranformation on the fly in the graph.  

```{r}
# going from gapminder to 
gapminder %>%
  group_by(continent,year) %>%
  summarize(mean_gdpPercap=mean(gdpPercap)) %>% 
  ggvis(~year, ~log(mean_gdpPercap), stroke = ~continent) %>% 
  layer_lines() %>% 
  add_axis("x", title = "Year", format='####') %>%
  add_axis("y", title = "Log GDP per capita per continent")
```

## 1.2 Tidyr - Wide to long and back again

### 1. Load the revenue.csv le into your R project.

```{r}
revenue <- read.csv('https://raw.githubusercontent.com/ucsdlib/win2016-gps-intro-R/gh-pages/data/revenue.csv', header = T)
```

### 2. By using gather(), `gather' the quarter revenues into Quarter and Revenue.
```{r}
longRev <- revenue %>% gather(Quarter, Revenue, Qtr.1:Qtr.4)
head(longRev)
```


### 3. By using separate(), `separate' each of the Quarter strings into Time Interval and Interval ID

```{r}
separateRev <- longRev %>% separate(Quarter, c("Time_Interval", "Interval_ID"))
head(separateRev)
```

### 4. By using unite(), re-unite the Time Interval and Interval ID and re-create the original Quarter variable.

```{r}
uniteRev <- separateRev %>% unite(Quarter, Time_Interval, Interval_ID, sep=".")
head(uniteRev)
```

###5. By using spread(), restore the original structure of the dataframe.

```{r}
wideRev <- uniteRev %>% spread(Quarter, Revenue)
head(wideRev)
```
