---
title: "r-homework1-edit"
author: "Reid Otsuji"
date: "1/11/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Due 1/17 - By Start of Class 6:30pm
## Instructions:
For this assignment you will be using a RMarkdown document. More info [here](http://rmarkdown.rstudio.com/). 
1. Type your code in the grey `code chunk` area.  
2. To run your code, click on the green triangle in the code chunk area.

## How to submit your homework:
1. Download the r-homework1.Rmd file in TritonEd and open in Rstudio.
2. Rename `r-homework1.Rmd` to `yourlastname_sudentID_r-homework1.Rmd` (save as in RStudio)
3. Type the answers in the uncommented lines (those without #) in the code chunks. Notice, for some lines, we have gotten you started with names of objects.
4. Save the Rmarkdown .Rmd file. 
5. Upload the .Rmd file to TritonEd (preferred) by 1/17 before 6:30pm. Or you can email your script directly to rotsuji@ucsd.edu. NOTE: Please use your UCSD email address when sending an email.  


### Exercise 1 : Making Vectors 1 (2.5 points)
a. create a vector composed of numbers from 1 to 10 using a R command  (hint: combine() c ()seq())
b. coerce the type of the vector to be string
c. Using a R command, find the first elements of the vector
d. Using a R command, find the first 5 elements of the vector
e. Using a R command, find the last element of the vector
f. Using a R command, find the last 5 elements of the vector
g. Using a R command, find the length of the vector
```{r eval=FALSE}
# Write your code below the # comment:
# a.



# b.


# c. 


# d.


# e. 



# f.


# g.  



```

### Exercise 2 : Making Vectors 2 (5 points)
Start by making a vector named 'myvector' with the numbers 1 through 26.
Create another vector named, 'myvectimestwo' by multiply the vector by 2, and give the resulting vector names A through Z (hint: there is a built in vector called `LETTERS`)
```{r eval=FALSE}
# Write code below:



```

---

### Exercise 3: Seeking Help (2.5 points)
1. Start by listing the preloaded datasets by typing `data()` in the console window.

2. Find the `mtcars` data set. Write the command to find the `help` information about the mtcars dataset.
```{r eval=FALSE}
# Write code below:



```


3. From the description, what does the dataset represent?



---

### Exercise 4: Factors (10 points)
Load the small cats dataset we created in class (feline-data.csv).  If you don't have this data set, see the Data Structures lesson <https://ucsdlib.github.io/win2018-gps-r/notes-edited/04-data-types.html>
```{r eval=FALSE}
# load the data set
cats <- read.csv(file="feline-data.csv") # edit the path in the read.csv function to the data file on your computer.
```

Try using `?read.csv` to figure out how to keep text columns (e.g. coats) as character vectors instead of factors; then write a command or two to show that the factor in `cats` is actually a `character vector` when loaded in this way.

```{r eval=FALSE}
# Complete these lines:
cats <- read.csv()
str()
```

---

### Exercise 5: Data Types and Structures (10 Points)
Create a list of length two containing a character vector for each of the sections in this part of the workshop:

* Data types
* Data structures

Populate each character vector with the **names of the data types and data structures**.
(Hint: Review data structures lesson) <https://ucsdlib.github.io/win2018-gps-r/notes-edited/04-data-types.html>

```{r eval=FALSE}
# complete the code
dataTypes <- c()  #character vector 
dataStructures <- c() #character vector 
answer <- list() #list of length two

```

---

### Exercise 6: Structure (10 points)
1. Use the function `data()` to display the list of Locate the builtin `iris` dataset. 

2. Write function used to investigate the structure of the dataset `iris`.

```{r eval=FALSE}
# Write code below:



```


3. Review the results of the last fucntion.  How many objects(obs) and variables are in the dataset? 





5. For the final step, list the commands you would use to find the `levels` attribute for the `species` varaiable. 
(Hint: start with assigning the iris database to a variable. Use `levels()` function.)
```{r eval=FALSE}
# Write code below:



```


---

### Exercise 7: Matrix (10 points)

1. Make a matrix called "mymatrix" containing the numbers 1:50, with 5 columns and 10 rows. Print the matrix out.
2. Make the "mymatrix" matrix above fill by row, not by column (its default behaviour). Print out the matrix.
(hint: read the `matrix` lesson section!)
 
```{r eval=FALSE}
# Write code below:



 
```
 
---



