---
title: "r-homework1-edit"
author: "Reid Otsuji"
date: "1/11/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Due 1/17 - By Start of Class 6:30pm
## How to submit your homework:
1. Download the [r-homework1.R](https://raw.githubusercontent.com/ucsdlib/win2018-gps-r/gh-pages/Homework/r-homework1.R) (Save As from your browser) script and open in Rstudio
2. Rename `r-homework1.R` to `yourlastname_sudentID_r-homework1.R` (save as in RStudio)
3. Type the answers in the uncommented lines (those without #) in the code chunks. Notice, for some lines, we have gotten you started with names of objects.
4. Save the script file. 
5. Upload the script file to TritonEd (preferred) by 1/17 before 6:30pm. Or you can email your script directly to rotsuji@ucsd.edu. NOTE: Please use your UCSD email address when sending an email.  


### Exercise 1 : Making Vectors (5 points)

Start by making a vector named 'myvector' with the numbers 1 through 26.
Create another vector named, 'myvectimestwo' by multiply the vector by 2, and give the resulting vector names A through Z (hint: there is a built in vector called `LETTERS`)
```{r eval=FALSE}
# Write code below:



```

---

### Exercise 2: Seeking Help (5 points)
1. Start by listing the preloaded datasets by typing `data()` in the console.

2. Find the `mtcars` data set. Write the command to find the `help` information about the mtcars dataset.
```{r eval=FALSE}
# Write code below:



```


3. From the description, what does the dataset represent?



---

### Exercise 3: Factors (10 points)
Load the small cats dataset we created in class (feline-data.csv).  If you don't have this data set, see the Data Structures lesson <https://ucsdlib.github.io/win2018-gps-r/notes-edited/04-data-types.html>
```{r eval=FALSE}
# load the data set
cats <- read.csv(file="feline-data.csv")
```

Try using `?read.csv` to figure out how to keep text columns (e.g. coats) as character vectors instead of factors; then write a command or two to show that the factor in `cats` is actually a `character vector` when loaded in this way.

```{r eval=FALSE}
# Complete these lines:
cats <- read.csv() 
str()
```

---

### Exercise 4: Data Types and Structures (10 Points)
Create a list of length two containing a character vector for each of the sections in this part of the workshop:

* Data types
* Data structures

Populate each character vector with the **names of the data types and data structures**.
(Hint: Review data structures lesson) <https://ucsdlib.github.io/win2018-gps-r/notes-edited/04-data-types.html>

```{r eval=FALSE}
# complete the code
dataTypes <- c()  #character vector 
dataStructures <- c() #character vector 
answer <- list() #list of length two

```

---

### Exercise 5: Structure (10 points)
1. Use the function `data()` to display the list of Locate the builtin `iris` dataset. 

2. Write function used to investigate the structure of the dataset `iris`.

```{r eval=FALSE}
# Write code below:



```


3. Review the results of the last fucntion.  How many objects(obs) and variables are in the dataset? 





5. For the final step, list the commands you would use to find the `levels` attribute for the `species` varaiable. 
(Hint: start with assigning the iris database to a variable. Use `levels()` function.)
```{r eval=FALSE}
# Write code below:



```


---

### Exercise 6: Matrix (10 points)

1. Make a matrix called "mymatrix" containing the numbers 1:50, with 5 columns and 10 rows. Print the matrix out.
2. Make the "mymatrix" matrix above fill by row, not by column (its default behaviour). Print out the matrix.
(hint: read the `matrix` lesson section!)
 
```{r eval=FALSE}
# Write code below:



 
```
 
---



