---
title: "Intro to R"
author: "Reid Otsuji adapted from Tim Dennis"
date: "01/04/2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

**55 min (40min teaching, 15min exercises)**
** should break 1/2 thru**
** shoot to finish at 3:15-25**

## Data Structures
##### Objectives 
* How to read data into R?
* What are the basic data types?
* How to represent categorical information in R?

* R deals with tabular data very well.
* Let's create a small data set for cats saved as a csv file.
* You can create `data/feline-data.csv` using a text editor (Nano), or within RStudio with the `File -> New File -> Text File` menu item.

```{r eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

##### Load csv in R
* Read the file in: 

```{r eval=FALSE}
cats <- read.csv(file = "data/feline-data.csv")
cats
```
* `read.csv` used for reading in tablular data stored in a text file, comma seperated values
* `tabs` and `Commas` are the most common punctuation characters used to seperate data points in csv files
* Tabs also common, use `read.delim`
* If not tabs or commas, you can use the more general `read.table`

* we can begin exploring the dataset using the `$` operator
* `$` operator pulls out columns by specifying them

```{r eval=FALSE}
cats$weight
```

```{r eval=FALSE}
cats$coat
```

* We can do operations on columns:
```{r eval=FALSE}
## say we discovered that the scale weighs two Kg light:
cats$weight + 2
```
```{r eval=FALSE}
paste("My cat is", cats$coat)
```

**but what about**
```{r eval=FALSE}
cats$weight + cats$coat
```

* understanding what happened here is **important** and key to sucessfully analyzing data in R

## Data Types 

* In the last example, if you guessed that `2.1 plus "black"` is nonsense, you are right
* important concept in programming language is `data types`
* we can ask what type of data something is in R for example:

```{r eval=FALSE}
typeof(cats$weight)
```

There are 5 main data types: 

* `double`  - double precision float number (used for math)
* `integer` - used for IDs or Values
* `complex` - 
* `logical` 
* `character`

```{r}
typeof(3.14)
```
```{r}
typeof(1L) # the L suffix forces the number to be an integer, since by default R uses float numbers, R stores as integer
```
**note** the L suffix for indicating an number is an iteger

```{r}
typeof(1+1i) #complex numbers with real and imaginary parts
```
```{r}
typeof(TRUE) # TRUE or FALSE
```

```{r}
typeof('banana')
```

* no matter how complicated our analysis is all data in R is interepeted as of one of these basic types

* as an example, a user added details of another cat
**edit data/feline-data and add new tabby data line `tabby, 2.3 or 2.4,1`
** re-save as `data/feline-data.csv`

```{r}
file.show("data/feline-data.csv")
```
* load the new file & check what type of data we find in `weight` column

```{r eval=FALSE}
cats <- read.csv(file="data/feline-data.csv")
typeof(cats$weight)
```

* oh no, our weights aren't the double type anymore!
* if we try to do the same math we did on them before we get unexpected results

```{r eval=FALSE}
cats$weight + 2
```

**what happened?**

* R reads a csv into a table, it insists that everything in the table is the same basic type
* if it can't understand in the column as a `double` then nobody in the column gets to be a double
* the **table** that R loaded our cats data into is called a **data.frame** and our first example of something called a **data structure**
   * a structure that R knows how to build out of basic data types.

* we can see that it is a `data.frame` by calling the `class` function on it:

```{r eval=FALSE}
class(cats)
```

* in order to use our data in R, we need to understand what the basic data sructures are & how they behave
* for now let's remove that extra line from out cats data and reload the data

* **open the file and delete the bottom line**

before: 
```{r eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
tabby,2.3 or 2.4,1
```
after:
```{r eval=FALSE}
coat,weight,likes_string
calico,2.1,1
black,5.0,0
tabby,3.2,1
```

```{r eval=FALSE}
cats <- read.csv(file="data/feline-data.csv")
```

## Vectors and Type Coercion

To better understand data structure behavior, let’s look at another of the data structures: the vector.

```{r}
my_vector <- vector(length = 3)
my_vector
```

* a `vector` is an `ordered list of things`, with a special condition that

*everything in the vector must be the same bacis data type* 

* if you dont choose a datatype, it'll default to `logcial`
* you can `declare an empty vector of whatever type you like`

```{r}
another_vector <- vector(mode='character', length = 3)
another_vector
```
* check the vector with str

```{r}
str(another_vector)
```

* the somewhat cryptic output from the `str` command indicates the basic data type found in this vector which is a`chr, character`
* also indicates the indexes of the vector [1:3]
* and a few examples of what's in the vector - empty strings here

for example using the cats dataset:

```{r eval=FALSE}
str(cats$weight)
```
* cats$weight is a vector, too-- the columns of data we load into R `data.frames` are all vectors
* and that's the root of why R forces everything in a coloumn to be the same basic  type

##### Why is R so opinionated about what we put in our columns of data? How does this help us?

By keeping everything in a column the same, we allow ourselves to make simple assumptions about our data; if you can interpret one entry in the column as a number, then you can interpret all of them as numbers, so we don’t have to check every time. This consistency, like consistently using the same separator in our data files, is what people mean when they talk about clean data; in the long run, strict consistency goes a long way to making our lives easier in R.

* you can make vectors with explicit contents with the`combine or c` function 

```{r}
concat_vector <- c(2,6,3)
concat_vector
```

* given what we've learned so far, what would this produce?:

```{r}
quiz_vector <- c(2,6,'3')
```


```{r}
str(quiz_vector)
```

* what happened above was *type conversion/coercion* and is the source of surprises and the reason we need to be aware of basic data types and how R deals with them and how R interprets data types
* When R encounters a mix of types (here in cats data, numeric and character) to be combined in to a vector, it will force them to be the same type

```{r}
coercion <- c('a', TRUE)
str(coercion)
```
```{r}
another_coercion_vector <- c(0, TRUE)
another_coercion_vector
```
* the coercion rules go: **logical -> integer -> numeric -> complex ->character**
  * Where `->` can be read as **are transformed into**
* You can try to force coercion against this flow using the `as.` functions

```{r}
character_vector_example <- c('0','2','4')
character_vector_example
```

```{r}
character_coerced_to_numeric <- as.numeric(character_vector_example)
character_coerced_to_numeric
```
```{r}
numeric_coerced_to_logical <- as.logical(character_coerced_to_numeric)
numeric_coerced_to_logical
```

* as you can see surprising things can happen when R forces one basic data type into another
* Nitty gritty of type coercion aside the point is:
  * if your data doesn't look like what you thought it was going to look like, `type coercion may well be to blame` 
* make sure everything is the same type in your vectors and your columns of `data.frames` or you will get bad surprises

* but coercion can be useful!
* ex. in our cat's data *likes_string* is number, but we know that the 1s and 0s actually represent `TRUE` and `FALSE` 
* we should use logical datatype here - whcih has 2 states - TRUE or FALSE 
* we can coerce this column by using `as.logical` function

```{r eval=FALSE}
cats$likes_string <- as.numeric(cats$likes_string)
str(cats$likes_string)
```

```{r eval=FALSE}
cats$likes_string <- as.logical(cats$likes_string)
cats$likes_string
str(cats$likes_string)
```

* the combine/concatenate `c()` function, will also append things to an existing vector:

```{r}

ab_vector <- c('a', 'b')
ab_vector
```

* example, Add a character 
```{r}
concat_example <- c(ab_vector, 'SWC')
concat_example
```

* you can also make a series of numbers

```{r}
mySeries <- 1:10
mySeries
```
```{r}
seq(10)
```

```{r}
seq(1,10, by=0.1)
```

* we can ask a few questions of vectors

```{r}
sequence_example <- seq(10)
head(sequence_example, n=2)
```
```{r}
tail(sequence_example, n=4)
```
```{r}
length(sequence_example)
```
```{r}
class(sequence_example)
```
```{r}
typeof(sequence_example)
```


* finally, you can give names to elements in your vector
```{r}
names_example <- 5:8
names(names_example) <- c("a", "b", "c", "d")
names_example
```

```{r}
names(names_example)
```

Challenge 1


## Data Frames

* we said that columns in `data.frames` were vectors

```{r eval=FALSE}
str(cats$weight)
```
```{r eval=FALSE}
str(cats$likes_string)
```
```{r eval=FALSE}
str(cats$coat)
```

## Factors

* another important data structure is a **factor**
* factors appear to be character data, but are used to represent categorical information
* Example: a vector of strings labelling cat colorations for all cats in our study:

```{r eval=FALSE}
coats <- c('tabby', 'tortoiseshell', 'tortoiseshell', 'black', 'tabby')
coats
```

```{r eval=FALSE}
str(coats)
```

* we can turn a vector into a factor like so

```{r eval=FALSE}
CATegories <- factor(coats)
```

```{r eval=FALSE}
class(CATegories)
```
```{r eval =FALSE}
str(CATegories)
```

* now R has noticed that there are three possible categories in our data
* but prints out a bunch of numbers instead
* R has replaced our human-readable categories with human readable `numbered indices`
* this is necessary as many statistical calculations utilise such numerical representations for categorical data:

```{r eval=FALSE}
typeof(coats)
```

```{r eval=FALSE}
typeof(CATegories)
```

Challenge 2


```{r eval=FALSE}
catsstring <- read.csv(file="data/feline-data.csv", stringsAsFactors=FALSE)

str(catsstring$coat)
```

```{r eval=FALSE}
catsstring <- read.csv(file="data/feline-data.csv", colClasses=c(NA, NA, "character"))
str(catsstring$coat)
```


* in modelling functions, it's important to know what the baseline levels are
* this is assumed to be the first factor, but by default `factors are labelled in aphabetical`
* you can change this by specifying levels:

```{r}
mydata <- c("case", "control", "control", "case")
factor_ordering_example <- factor(mydata, levels = c("control", "case"))
str(factor_ordering_example)
```

* in this case we explicitly told R that 'control' should be represented by 1 and case by 2
* this designation can be very important for interpreting the results of stats model

## Lists

* another data structure you'll want to learn is a `list`
* a list is simpler in some ways than other types because you can put anything in it

```{r}
list_example <- list(1, "a", TRUE, 1+4i)
list_example
```
```{r}
another_list <- list(title = "Research Bazaar", numbers = 1:10, data = TRUE )
another_list
```

* we can now understand something surprising in our cats `data.frame` what happens if we:

```{r eval=FALSE}
typeof(cats)
```

* we see that `data.frames` look like `lists` under the hood 
* this is because a data.frame is really a list of vectors and factors 
* in order to hold those columns that are a mix of vectors and factors, the data.frame needs something a bit more flexible than a vector to put all those columns together into a table
* `data.frame` is a special ist in which all vectors must have same length

* in our cats exampel we have an integer, a double and logical variable

```{r eval=FALSE}
cats$coat
```
```{r eval=FALSE}
cats[,1]
```
```{r eval=FALSE}
typeof(cats[,1])
```
```{r eval=FALSE}
str(cats[,1])
```

* each row is an `observation` of different variables, itself a `data.frame` and thus can be composed of element of different types

```{r eval=FALSE}
cats[1,]
```
```{r eval=FALSE}
typeof(cats[1,])
```
```{r eval=FALSE}
str(cats[1,])
```


Challenge 3



## Matrices 

* Last but not least is the matrix
* we can declare a matrix of zeros:

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
matrix_example
```
* And similar to other data structures, we can ask things about our matrix:

```{r}
class(matrix_example)
```
```{r}
typeof(matrix_example)
```
```{r}
str(matrix_example)
```
```{r}
dim(matrix_example)
```
```{r}
nrow(matrix_example)
```


Challege 4


```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
length(matrix_example)
```




Challenge 5


```{r}
x <- matrix(1:50, ncol=5, nrow=10)
```


```{r}
x <- matrix(1:50, ncol=5, nrow=10, byrow = TRUE) # to fill by row
```

