---
title: "intro R - Exploring Data Frames"
author: "Reid Otsuji adapted from Tim Dennis"
date: "01/4/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Lesson objective:
* Be able to add and remove rows and columns.
* Be able to remove rows with NA values.
* Be able to append two data frames
* Be able to articulate what a factor is and how to convert between factor and character.
* Be able to find basic properties of a data frames including size, class or type of the columns, names, and first few rows.

**How can I manipulate a dataframe?**

* At this point, you’ve see it all - in the last lesson, we toured all the basic data types and data structures in R. 
* Everything you do will be a manipulation of those tools. 
* most of the time using R the star of show is the `data.frame` - the table created by loading data from a csv file

Now we will learn a few more things about working with data frames

### adding columns and rows in a data.frame
* we learned that columns in data.frames were vecotrs, so that we have  type of data down a column


*reload cats csv if needed to reset cats variable*
```{r}
cats <- read.csv(file="feline-data.csv")
```
* if we want a new column, we start by making a vector

```{r}
age <- c(2,3,5,12)
cats
```

* we can then add a column via: 

`cbind(cats, age)`

* produces error
Error in data.frame(..., check.names = FALSE): arguments imply differing number of rows: 3, 4

`cats <- cbind(cats, age)`
* produces error
Error in data.frame(..., check.names = FALSE): arguments imply differing number of rows: 3, 2

* what happened? 
*  Of course, R wants to see one element in our new column for every row in the table:
```{r}
nrow(cats)
```
```{r}
length(age)
```

* So for it to work we have to have `nrow(cats) = length(age)`. 
* Let’s store it into cats and overwite the contents of that data frame.


```{r}
cats
```

```{r}
age <- c(4,5,8)
cats <- cbind(cats, age)
```


```{r}
cats
```

Now how about adding rows - in this case, we saw last time that the rows of a data.frame are made of lists:

```{r}
newRow <- list("tortoiseshell", 3.3, TRUE, 9)
cats <- rbind(cats, newRow)
```

#### Factors
* another thing to look out for - when R creates a factor, it only allows whatever is originally there when our data was loaded which was ('black', 'calico', 'tabby')
* anything that doesn't fit into one of its categories is rejected and becomes NA
* to add the new row, we need to explicitly add is as a level in the factor

```{r}
levels(cats$coat)
```

```{r}
levels(cats$coat) <- c(levels(cats$coat), 'tortoiseshell')
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9))
```
```{r}
cats
```

* Alternatively we can change a factor column to a character vector
* but we lose handy categories of the factor but can subseqently add any word we want to the column 

```{r}
str(cats)
```

```{r}
cats$coat <- as.character(cats$coat)
```

```{r}
str(cats)
```


## Removing rows
* we now know how to add rows and columns to our data.fram in R 
* but in our work so far we've accidentlly added a garbage row

```{r}
cats
```
* we can ask for a data.fram minus this offending row: 
```{r}
cats[-4,]
```

* notice the comma with nothing after it to indicate we want to drop that row

* **Note** we can remove both new rows at once by putting the row numbers inside of a vector: `cats[c(-4,-5),]`

* or we can also drop all rows with NA values

```{r}
na.omit(cats)
```

* let's reassign the output to cats, so our canges are made permanent

```{r}
cats <- na.omit(cats)
```

##appending to a data frame

* the key to remember when adding data to a data.frame is that `columns` are **vectors or factors, and rows are lists**
* we can also glue two data.frames together with rbind:

```{r}
cats <- rbind(cats, cats)
cats
```

* but now row names are complicated 
* we can remove the rownames and R will auto re-name them sequentially

```{r}
rownames(cats) <- NULL
cats
```

Challenge 1

## Realistic example
**So far:**

* basics of manipulating `data.frames` with our cat data
* let's work on a more realistic data set
* we're going to read in the gapminder dataset
*make sure everyone has the data saved to desktop and in a RStudio working directory*

#### Gapmidner data info:
*  For each of 142 countries, the package provides values for life expectancy, GDP per capita,
and population, every five years, from 1952 to 2007
* The main data frame gapminder has 1704 rows and 6 variables:
* `country` - factor with 142 levels
* `continent` - factor with 5 levels
* `year` - ranges from 1952 to 2007 in increments of 5 years
* `lifeExp` - life expectancy at birth, in years
* `pop` - population
* `gdpPercap` - GDP per capita (US$, inflation-adjusted)

```{r}
gapminder <- read.csv("gapminder-FiveYearData.csv") #file path will be different on your computer. Us the path where the .csv files is saved on your computer
```

#### loading data tips
* mention tab-separated values files (.tsv)
* to specify a tab separator use "\\tab" or `read.delim()`
* files can be downloading via the internet using the `download.file` and the read.csv can be executed to read the downloaded file such as: 

```{r eval=FALSE}
download.file("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv", destfile = "data/gapminder-FiveYearData.csv")
gapminder <- read.csv("data/gapminder-FiveYearData.csv")
```

* alternately you can read files directly into R from the internet by replacing file paths with a web address in `read.csv`
* **note** when you do this there is not **local copy** saved

```{r eval=FALSE}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")
```

* You can read directly from excel spreadsheets without converting them to plain text by using the readxl package 

#### working with gapminder dataset
* let's investigate gapminder
* the first thing we should always do is check out what the data looks like with `str`
```{r}
str(gapminder)
```
```{r}
typeof(gapminder$year)
```
```{r}
typeof(gapminder$lifeExp)
```
```{r}
typeof(gapminder$country)
```
```{r}
str(gapminder$country)

```

* we can also interrogat the data.frame for info about its dimensions
* remember that str(gapminder) said there were 1704 observations of 6 variables in gapminder
* what do you think the following will produce?

```{r}
length(gapminder)
```

* a fair guess would say that the length of a data.frame would be the number of rows it has (1704)
* not the case, remember, a data.frame is a *list of vectors and factors*:

```{r}
typeof(gapminder)
```

* when length gave us 6 it's because gapminder is built out of a list of 6 columns 
* to get number of rows and columns in our data set try:

```{r}
nrow(gapminder)
```
```{r}
ncol(gapminder)
```

* or both at once

```{r}
dim(gapminder)
```

* we also would want to know what the titles of all the columns are, so we can ask for them by name:

```{r}
colnames(gapminder)
```
* it is important to ask if the structure R is reporting matches our intuition or expectations
* do the basic data types reported make sense?
* if not we need to sort out problems now before they turn into negative surprises down the road
* remmeber how R interprets data, and the importance of strict consistency in how we record our data

* once we are happy that the data types and structures seem reasonable, it's time to start digging into our data properly

```{r}
head(gapminder)
```
* to make sure our analysis is reproducible we should put the code into a script file so we can come back later


Challenge 2

Challege 3

## Cheat Sheets
* mention cheatsheets that are avialble
<https://www.rstudio.com/resources/cheatsheets/>



