---
title: "Subsetting Data"
author: "Reid Otsuji adapted from Tim Dennis"
date: "01/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* To be able to subset vectors, factors, matrices, lists, and data frames
* To be able to extract individual and multiple elements: by index, by name, using comparison operations
* To be able to skip and remove elements from various data structures.

## Subsetting Data

* There are Six different ways we can subset any kind of object
* Three different subsetting operators for the different data structures
* let's start with simple numeric vectors

```{r}
x <- c(5.4, 6.2, 7.1, 4.8, 7.5)
names(x) <- c('a', 'b', 'c', 'd', 'e')
x
```

* simple vectors that cannot be further simplified that contain character strings, numbers, or logical values are called `atomic vectors`


## Accessing elements using their indices

* extract elements of a vector we can give their corresp. indes
```{r}
x[1]
x[4]
```

* it may not look like it but the square brackets operator is a function
* it means: "get me the nth element"
* ask for multiple elements at once

```{r}
x[c(1, 3)]
```

* or slices of the vector

```{r}
x[1:4]
```

* the `:` operator creates a sequence of numbers from the left element to the right

```{r}
1:4
c(1, 2, 3, 4)
```

* we can ask for the same element multiple times

```{r}
x[c(1,1,3)]
```
* if we ask for a number outside the vecotr, R will return missing values:

```{r}
x[6]
```

* this is a vector of length of one containing an NA, whose name is also NA
* if we ask for the 0th element, we get an empty vector:

```{r}
x[0]
```

NOTE about vector numbering:
* R indexing starts at 1 - many programming languages (c & python) indexing starts at 0

## Skipping and removing elements

* if we use a negative number as the index of a vector, R will return every element EXCEPT for the one specified

```{r}
x[-2]
```

* we can skip multiple elements

```{r}
x[c(-1,-5)] #or x[-c(1,5)]
```

TIP for Order of Operations:
* a common trip up for novices occurs when trying to skip slices of a vector
* most ppl first try to negate a sequence like so:

```{r eval=FALSE}
x[-1:3]
```
This returns a cryptic error: 

```{r eval=FALSE}
Error in x[-1:3]: only 0's may be mixed with negative subscripts
```


* remember order of operations. 
* `:` is really a function
  * what happens is it takes first arguemnt as `-1`, and a second as `3`, so getnerates a sequence of numbers: `c(-1, 0, 1, 2, 3)`

* correct solution is to wrap that function call in brackets, so that the `-` operator applies tot he result

```{r}
x[-(1:3)]
```

* to remove elements from a vector we need to assign the results back onto the object

```{r}
x <- x[-4]
x
```
#### challenge 1

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-1>

## Subsetting by name

* extract elements by using their name, instead of extracting by index:
```{r}
x[c("a","c")]
```

    * usuall more reliable way to subset objects, index changes more than names
    * unfortunatelly can't skip or remove as easily
    
## Subsetting through other logical operations

* Also use any logical vector to subset:
```{r}
x[c(FALSE,FALSE,TRUE,FALSE,TRUE)]
```
* Comparison operators (e.g. `>`, `<`, `==`) are logical vectors we can use the to subset vectors
* This example statement gives the same result as the previous one

```{r}
x[x >7]

```
* breaking it down, the statement first evaluates `x > 7`
* generates a logical vector `c(FALSE, FALSE, TRUE, FALSE, TRUE)` then selects the elements of `x` corresponding to the `TRUE` values

* next example, we can use `==` to mimic the previous method of indexing by name
* remember you have to use `==` rather than `=` for comparisons
```{r}
x[names(x) == "a"]
```

** TIP: Subsetting through `other` logical operators**
There are many situations in which you will wish to combine multiple conditions. To do so several logical operations exist in R:

* `|` logical `OR`: returns `TRUE`, if either the left or right are `TRUE`.
* `&` logical `AND`: returns `TRUE` if both the left and right are `TRUE`
* `!` logical `NOT`: converts `TRUE` to `FALSE` and `FALSE` to `TRUE`
* `&&` and `||` compare the individual elements of two vectors. Recycling rules also apply here.



#### challenge 3 
<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-1>

** TIP: non-unique names ** 
* be aware that it is possible for multiple elements in a vector to have the same name. 
* e.g. for a data frame, columns can have the same name. (but R tries to avoid this and row names must be unique).

For example:

```{r}
x <- 1:3
x
```
```{r}
names(x) <- c('a','a','a')
x
```
```{r}
x['a'] # only returns first value
```
```{r}
x[names(x) == 'a'] # returns all three values
```

## Skipping named elements
* skipping of removing named elemnts is a little harder 
* If we try to skip one named element by negating the string, R complains (slightly obscurely) that it doesn’t know how to take the negative of a string:

```{r eval=FALSE}
x <- c(a=5.4, b=6.2, c=7.1, d=4.8, e=7.5) # we start again by naming a vector 'on the fly'
x[-"a"]
```

* However, we can use the `!=` (not-equals) operator to construct a logical vector that will do what we want:

```{r}
x[names(x) != "a"]
```

* Skipping multiple named indices is a little bit harder still
* Suppose we want to drop the `"a"` and `"c"` elements, so we try this:
```{r}
x[names(x)!=c("a","c")]

```

* R did something, but it gave us a warning that we ought to pay attention to - and it apparently gave us the wrong answer (the "c" element is still included in the vector)!

* So what does `!=` actually do in this case? That’s an excellent question.


## challenge 2 
<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting/>

## Recycling
* Let’s take a look at the comparison component of this code:

```{r}
names(x) != c("a", "c")
```
* Why does R give FALSE as the third element of this vector, when `names(x)[3] != "c"` is obviously false? 

* When you use `!=`, R tries to compare each element of the left argument with the corresponding element of its right argument. 

* What happens when you compare vectors of different lengths? 
**Show graphic Example from lesson**

** Explain Graphics ** 
* In this case R repeats `c("a", "c")` as many times as necessary to match `names(x)`, i.e. we get `c("a","c","a","c","a")`.

* Since the recycled `"a"` doesn’t match the third element of `names(x)`, the value of `!=` is `FALSE`.

* Because in this case the `longer vector length (5)` isn’t a multiple of the shorter vector length (2)`, R printed a warning message.

* If we had been unlucky and `names(x)` had contained six elements, R would silently have done the wrong thing (i.e., not what we intended it to do). 

* This recycling rule can can introduce hard-to-find and subtle bugs!

** how to handle recycling **
* The way to get R to do what we really want (match each element of the left argument with all of the elements of the right argument) it to use the `%in%` operator.

* Getting help for operators; 
    - `help("%in%")` or `?"%in%"`
    
* The `%in%` operator goes through each element of its left argument, in this case the names of `x`, and asks, “Does this element occur in the second argument?”.

* Here, since we want to exclude values, we also need a `!` operator to change `“in”` to `“not in”`:
```{r}
x[! names(x) %in% c("a","c")]
```
 

**R recycling rule example same as lesson graphics** 

```{r eval=FALSE}
names(x) == c('a', 'c') #warnings 
#== works slightly differently than %in%. It will compare each element of its left argument to the corresponding element of its right argument.
# R recycles the shorter vector in a equality comparison

c("a", "b", "c", "e")  # names of x
   |    |    |    |    # The elements == is comparing
c("a", "c")

c("a", "b", "c", "e")  # names of x
   |    |    |    |    
c("a", "c", "a", "c") # The elements == is comparing
```




**Handling special values**

At some point you will encounter functions in R which cannot handle missing, infinite, or undefined data.

special functions to deal with this: 

* `is.na` will return all positions in a vector, matrix, or data.frame containing `NA`.
* likewise, `is.nan`, and `is.infinite` will do the same for `NaN` and `Inf`. 
* `is.finite` will return all positions in a vector, matrix, or data.frame that do not contain `NA`, `NaN` or `Inf`.
* `na.omit` will filter out all missing values from a vector

**factor subsetting**

Factor subsetting works the same way as vector subsetting.

```{r}
f <- factor(c("a", "a", "b", "c", "c", "d"))
f[f == "a"]
```

```{r}
f[f %in% c("b", "c")]
```

```{r}
f[1:3]

```
* An important note is that skipping elements will not remove the level even if no more of that category exists in the factor:
```{r}
f[-3]
```



**Matrix subsetting**

Matrices are also subsetted using the `[` function. In this case it takes two arguments: the first applying to the rows, the second to its columns:

```{r}
set.seed(1)
m <- matrix(rnorm(6*4), ncol=4, nrow=6)
m[3:4, c(3,1)]
```
* You can leave the first or second arguments blank to retrieve all the rows or columns respectively:

```{r}
m[, c(3,4)] #column
```

* If we only access one row or column, R will automatically convert the result to a vector:
```{r}
m[3,] #if grabbing 1 row, R will convert to vector
```

* If you want to keep the output as a matrix, you need to specify a third argument; `drop = FALSE`:

```{r}
m[3, , drop=FALSE] # keep as a matrix specify a third argument; drop = FALSE:
```


* Unlike vectors, if we try to access a row or column outside of the matrix, R will throw an error:
```{r eval=FALSE}
m[, c(3,6)] #will throw error if out of range
```

* because matricies are really vectors, you can just use single indexing
```{r}
m[5] #not very useful
```

* This usually isn’t useful, and often confusing to read.

* but matricies are populated by `column-major format` by default and elements are arraged column wise

```{r}
matrix(1:6, nrow=2, ncol=3)
```

* If you wish to populate the matrix by row, use `byrow=TRUE`:

```{r}
matrix(1:6, nrow=2, ncol=3, byrow=TRUE) #populate by row use  byrow=TRUE

```

* Matrices can also be subsetted using their rownames and column names instead of their row and column indices.

## challenge 4

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting#challenge-4>

## List subsetting
* Now we’ll introduce 3 functions to subset lists.
* Three functions to subset lists are `[`, `[[`, `$`

* Using `[` will always return a list. If you want to subset a list, but not extract an element, then you will likely use `[`.

```{r}
xlist <- list(a = "Global Policy", b = 1:10, data = head(iris))
xlist[1]
```
* this is a list with 1 element

* we can subset elements teh same way as atomic vector using `[`
```{r}
xlist[1:2]
```


* to get at **individual elements** of a list you need to use `[[`
```{r}
xlist[[1]]
```
* notice the result is a vector, not a list

* You can't extract more than one element at once:
```{r eval=FALSE}
xlist[[1:2]]
```

```{r eval=FALSE}
xlist[[-1]]
```
* and you can't use it to skil elements
    
* but you can use names to both subset and extract elements
```{r}
xlist[['a']]
```

* the function `$` is a shorthand way for extracting elements by name:
```{r}
xlist$data
```

## challenge 5

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-5>

## Data frames

* Remember data frames are lists under the hood, so similar methods apply.
* `[` with one argument will act the same was as for lists, where each list element corresponds to a column. The resulting object will be a data frame:

```{r}
gapminder <- read.csv(file="data/gapminder-FiveYearData.csv")
head(gapminder[3])
```

Similarly, `[[` will act to extract a single column:
```{r}
head(gapminder[["lifeExp"]]) # [[ will act to extract a single column
```

* And `$` provides a convenient shorthand to extract columns by name:
```{r}
head(gapminder$year) #$ provides shorthand to extract columns by name
```

* With two arguments, `[` behaves the same way as for matrices:
```{r}
gapminder[1:3,]
```

* If we subset a single row, the result will be a data frame (because the elements are mixed types):
```{r}
gapminder[3,] #is a data frame b/c of the mixed types

```
## Challenge 7 

<http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting##challenge-7>








