---
title: "Functions"
author: "Reid Otsuji adapted from Tim Dennis"
date: "01/19/2018"
output: html_document
---

Goal: To be to define a function that takes arguments and return a value from a function. Overall, 
be able to explain why we should divide programs into small, single-purpose functions.


##What is a function?

Functions gather a sequence of operations into a whole, preserving it for ongoing use. Functions provide:

* a name we can remember and invoke it by
* relief from the need to remember the individual operations
* a defined set of inputs and expected outputs
* rich connections to the larger programming environment

As the basic building block of most programming languages, user-defined functions constitute **programming** as much as any single abstraction can. If you have written a function, you are a computer programmer.

## Creating a function 

Open a R Script `File > New File > R Script`:

```{r}
my_sum <- function(a, b) {
  the_sum <- a + b
  return(the_sum)
}
```

Using (calling) this function: 

```{r}
my_sum(10, 15)

```

Components of a function: 

1. body 
2. arguments
3. environment

```{r}
body(my_sum)
```
```{r}
formals(my_sum)
```
```{r}
#args(my_sum) #for a more human readable version
environment(my_sum)

```
** show environment window - read only functions view ** 

Let's define a function that converts Fahrenheit to Kelvin.

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

* define function fahr_to_kelvin by assigning it to the output of `function`. 
* List of arguments (e.g.temp) in the parenthese
* The `body` of the funtion -- the operations that are executed -- are within the curly braces `{}`
* indent statements 2 spaces inside body -- makes code easier to read

**Calling the function** 

* when we call the function, we pass its values to the arguments in the parentheses, so they can be used in the body (function)
* we use a return statement to send back results of the call - `Return Statement` is a statement that causes a function to stop executing and return a value back to the call immediately.
* return state optional in R - used for clarifying the last variable on the last line in the body.

```{r}
#freezing point of water
fahr_to_kelvin(32)

#boiling point of water
fahr_to_kelvin(212)
```


### Challenge 1
<http://swcarpentry.github.io/r-novice-gapminder/10-functions#challenge-1>

Write a function called `kelvin_to_celsius` that takes a temperature in Kelvin and returns that temperature in Celsiu. Hint: To convert from Kelvin to Celsius you minus 273.15
```{r}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}

```


## Combining functions

* The real power of functions comes when we mix, match and combine them into ever larger chunks to get what we need

* As an example, let's define two functions that will convert temperature from Fahrenheit to Kelvin, and Kelvin to Celsius

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}

kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

* to perform this calculation in one line of code, by “chaining” functions together, we can run:

```{r}
#freezing point of water in Celsius
kelvin_to_celsius(fahr_to_kelvin(32.0))
```



### Challenge 2

<http://swcarpentry.github.io/r-novice-gapminder/10-functions#challenge-2>

## More on combining functions

### Calculate GDP from GapMinder dataset
Now, we’re going to define a function that calculates the Gross Domestic Product of a nation from the data available in our dataset:

* make sure dataset is loaded as 'gapminder'

```{r}
gapminder <- read.csv(
  file = 'data/gapminder-FiveYearData.csv',
  header = T
  
)
head(gapminder)
```

* type function example:

```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat){
  gdp <- dat$pop * dat$gdpPercap
  return(gdp)
}

```

* defined calcGDP by using function.
* List arguments `dat` in parentheses
* The body of the function calculates `gdp`
* `return statment` to stop executing and send back the result.

```{r}

calcGDP(head(gapminder))

```
* When we call the function `calcGDP()`, the values  we pass to it `head(gapminder)` are assigned to the arguments, which become variables inside the body of the function.


* output not very informative. let's add country and year to our arguments to extract year and country

```{r}
# Takes a dataset and multiplies the population column
# with the GDP per capita column.
calcGDP <- function(dat, year=NULL, country=NULL) {
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
  gdp <- dat$pop * dat$gdpPercap

  new <- cbind(dat, gdp=gdp)
  return(new)
}

```

* let's save as a script so we can load our fuctions and source them.

* you can load in the functions into our R session by using the source function

```{r eval=F}
source("functions/functions.R")
```

Lot going on in this function: 

* function now subsets the data by `country` and/or `year` if their arguments aren't empty. 
* then calculates the `gdp` on the resulting subsetted data
* function then adds the result to a new column to the subsetted data and returns result.

* GDP for year
```{r}
head(calcGDP(gapminder, year=2007))

```

* GDP for specific country
```{r}
calcGDP(gapminder, country="Australia")

```

* or both year and country

```{r}
calcGDP(gapminder, year=2007, country="Australia")

```

* walk through body of function:
* added two arguments `country` and `year` with default value of `null` meaning arguments will take on vaules unless specified
```{r eval=FALSE}
calcGDP <- function(dat, year=NULL, country=NULL) {
```

* we check to see if they are not null with the if operator 
```{r eval=FALSE}
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
```
* this was done so function can be more flexible for calculating GDP for whole data set, single year, single country, year and country (like examples)

* and then subset the rows

* scoping - any variables or functions created or modified in body of the function only exist for the lifetime of the functions execution. 
* E.g. variables `dat`, `gdp`, and `new` only exist in the body of the function. 
```{r eval=FALSE}
gdp <- dat$pop * dat$gdpPercap
  new <- cbind(dat, gdp=gdp)
  return(new)
}
```

* lastly, calculate gdp and creeate a new data frame with column to return (output) 

*TIP: Important to test and document your functions for others to understand, how to use and to make sure the function does what you think it will do.

